Verification of OpenSSL's SHA3 \keccak\ function using the Software Analysis Workbench

### 2022-01-04 paper/fips202.tex
Revise the while-method and text;2C

\newsavebox{\PiCry}
\begin{lrbox}{\PiCry}
  \begin{lstlisting}[language=Cryptol, numbers=left, escapeinside={;}{;}]
    type STATE_ARR = [5][5][64];\label{line:state};
    LIST4 = [0..4];\label{line:list4};
    LIST63 = [0..63];\label{line:list63};
    pi : STATE_ARR -> STATE_ARR
    pi a = [  [  [a @x @((x + 3*y) % 5) @z;\label{line:pi};
                 | z <- LIST63] 
              | y <- LIST4] 
           | x <- LIST4]
  \end{lstlisting}
\end{lrbox}

\newsavebox{\PiCryLib}
\begin{lrbox}{\PiCryLib}
  \begin{lstlisting}[language=Cryptol]
    pi a = for LIST4 (&x ->
             for LIST4 (&y -> 
               for LIST63 (&z -> 
                 a @x @((x + 3*y) % 5) @z)))
  \end{lstlisting}
\end{lrbox}

@@

\begin{figure}[t]
  \begin{center}
    \begin{tabular}{l}
      \usebox{\fipsPi} \\ \\
      \multicolumn{1}{c}{(a)} \\ \\
      \usebox{\PiCry} \\ \\
      \multicolumn{1}{c}{(b)} \\ \\
      \usebox{\formethod} \\ \\
      \multicolumn{1}{c}{(d)} \\ \\ 
      \usebox{\PiCryLib} \\ \\
      \multicolumn{1}{c}{(c)}
    \end{tabular}
  \end{center}
  \caption{The $\pi$ function. (a) The \fips\ specification. (b) The \cryptol\ with comprehensions. (c) The for-method definition. (d) The \cryptol\ with the for-method.}
  \label{fig:pi}
\end{figure}

@@

The \cryptol\ definition using list comprehensions is given in \figref{fig:pi}(b).
\lineref{line:state}) defines the state as a $5 \times 5$ array of 64-bit words.
\lineref{line:list4} and \lineref{line:list63} are the lists for the list comprehensions.
These are the indices that are used in the computation and correspond to the finite domains for $x$, $y$, and $z$ defined in the quantification in \figref{fig:pi}(a).
\lineref{line:pi} starts the definition of $\pi$. 
Here the comprehensions are nested with the list indices appearing in reverse order to follow the nesting.
Although the list comprehensions are not unreadable, the \cryptol\ can be improved by creating a for-method to hide some of the details.

\figref{fig:pi}(c) is the \cryptol\ definition of the for-method.
It takes two arguments: a list, \emph{vals}, for the comprehension, and a function, $f$, to apply to each element of the list.
Its definition is as expected: it applies $f$ to each element in \emph{vals}.

@@

\subsection{The while-method}
Another common phrasing is the \emph{let...for...let}. 
This defines a state upon which a folding occurs, iteratively condensing the information into a final state. 
Imperatively, it is most comparable to a while loop. 
As shown by Figure \ref{fig:nistRC}, is present in the rc function defined by the NIST specification.

\begin{figure}[h]
  \centering
\begin{lstlisting}[basewidth = {.5em},basicstyle={\scriptsize}]
1. If t mod 255 = 0, return 1.
2. Let R = 10000000.
3. For i from 1 to t mod 255, let:
  a. R = 0 || R;
  b. R[0] = R[0] ^ R[8];
  c. R[4] = R[4] ^ R[8];
  d. R[5] = R[5] ^ R[8];
  e. R[6] = R[6] ^ R[8];
  f. R = Trunc8[R].
4. Return R[0].
\end{lstlisting}
\caption{NIST RC definition}
\label{fig:nistRC}
\end{figure}

Maintaining this state through Cryptol's list comprehensions can be difficult.
The comprehensions require a list to iterate through. 
Since the \emph{for} utility function is simply a wrapper for list comprehensions, it is equally unsuited for visual verification on such steps. 
The number of iterations in the RC function is bounded by [1, 254] by its modular condition.
This range provides the list to iterate over as required by Cryptol list comprehensions. 
From these observations, a rather ungainly Cryptol definition results.
This definition is shown in Figure \ref{fig:cryptolRC}.

\begin{figure}[h]
  \centering
\begin{lstlisting}[language=Cryptol]
rc : [64] -> Bit
rc t = if (t % 255) == 0 
  then True 
  else rs !0 @0 where
    rs = [128:[8]] # [
      if i <= t % 255
        then (take'{8} ([
          (r@0) ^ (r@8),
          r@1,
          r@2,
          r@3, 
          (r@4) ^ (r@8),
          (r@5) ^ (r@8),
          (r@6) ^ (r@8),
          r@7,
          r@8])
            where r = [False] # (rs @(i - 1)))
        else rs @(i - 1)
      | i <- [1..254]]
\end{lstlisting}
\caption{Cryptol RC definition}
\label{fig:cryptolRC}
\end{figure}

Not only is this definition computationally slower as it runs 255 loops every time it is called, it also awkwardly indexes into a list of previous states. 
This differs vastly from the NIST specificiation both visually and functionally. 
The \emph{while} utility function, accepts an initial state, condition function, and body function to recursively loop. 
It consolodates the information of the previous state until the condition is sastified. 
This definition is found in Figure \ref{fig:cryptolamendedRC}.

\begin{figure}[h]
  \centering
\begin{lstlisting}[language=Cryptol]
rc : [64] -> Bit
rc t = if (t % 255) == 0 
  then True 
  else (while {i = 1, R = 128:[8]}
    (&state -> state.i <= t % 255)
    (&state -> {
      i = state.i + 1, 
      R = (take'{8} ([
        (r@0) ^ (r@8),
        r@1,
        r@2,
        r@3,
        (r@4) ^ (r@8),
        (r@5) ^ (r@8),
        (r@6) ^ (r@8),
        r@7, 
        r@8]) 
          where r = [False] # state.R)})
    ).R @0
\end{lstlisting}
\caption{Amended Cryptol RC definition}
\label{fig:cryptolamendedRC}
\end{figure}

The final state is returned by the \emph{while} function. 
This iterates only until the condition is met, avoids the clutter of indexing within a larger array of all previous states, and places the condition in a more obvious location before the loop body. 
This utility function is defined in Figure \ref{fig:cryptolWhile}.

\begin{figure}[h]
  \centering
\begin{lstlisting}[language=Cryptol]
while : {a} a -> (a -> Bit) -> (a -> a) -> a
while state cond f = if (cond state)
  then (while (f state) cond f)	
  else state
\end{lstlisting}
\caption{Cryptol While utility function}
\label{fig:cryptolWhile}
\end{figure}

>>>>

\newsavebox{\PiCry}
\begin{lrbox}{\PiCry}
  \begin{lstlisting}[language=Cryptol, numbers=left, escapeinside={;}{;}]
    type STATE_ARR = [5][5][64];\label{line:state};
    pi : STATE_ARR -> STATE_ARR
    pi a = [  [  [a @x @((x + 3*y) % 5) @z;\label{line:pi};
                 | z <- [0..63]];\label{line:list};
              | y <- [0..4]] 
           | x <- [0..4]]
  \end{lstlisting}
\end{lrbox}

\newsavebox{\PiCryLib}
\begin{lrbox}{\PiCryLib}
  \begin{lstlisting}[language=Cryptol]
    pi a = for [0..4] (&x ->
             for [0..4] (&y -> 
               for [0..63] (&z -> 
                 a @x @((x + 3*y) % 5) @z)))
  \end{lstlisting}
\end{lrbox}

@@

\begin{figure}[t]
  \begin{center}
    \begin{tabular}{l}
      \usebox{\fipsPi} \\ \\
      \multicolumn{1}{c}{(a)} \\ \\
      \usebox{\PiCry} \\ \\
      \multicolumn{1}{c}{(b)} \\ \\
      \usebox{\formethod} \\ \\
      \multicolumn{1}{c}{(c)} \\ \\ 
      \usebox{\PiCryLib} \\ \\
      \multicolumn{1}{c}{(d)}
    \end{tabular}
  \end{center}
  \caption{The $\pi$ function. (a) The \fips\ specification. (b) The \cryptol\ with comprehensions. (c) The for-method definition. (d) The \cryptol\ with the for-method.}
  \label{fig:pi}
\end{figure}

@@

The \cryptol\ definition using list comprehensions is given in \figref{fig:pi}(b).
\lineref{line:state}) defines the state as a $5 \times 5$ array of 64-bit words.
\lineref{line:pi} starts the definition of $\pi$. 
Here the comprehensions are nested with the list indices appearing in reverse order to follow the nesting.
\lineref{line:list} gives the domain for $z$ as a list with values from 0 to 63.
The domains for $x$ and $y$ are given similarly.
The domains for $x$, $y$, and $z$ all match the domains defined in the quantification in \figref{fig:pi}(a).
Although the list comprehensions are not unreadable, the \cryptol\ can be improved by creating a for-method to hide some of the details.

\figref{fig:pi}(c) is the \cryptol\ definition of the for-method.
It takes two arguments: a list, \emph{vals}, for the comprehension, and a function, $f$, to apply to each element of the list.
Its definition is as expected: it creates a new list that is the result of applying $f$ to each element in \emph{vals}.

@@

\subsection{The while-method}

\newsavebox{\fipsRc}
\begin{lrbox}{\fipsRc}
  \begin{lstlisting}[basewidth = {.5em},basicstyle={\small}]
    1. If t mod 255 = 0, return 1.
    2. Let R = 10000000.
    3. For i from 1 to t mod 255, let:
      a. R = 0 || R;
      b. R[0] = R[0] ^ R[8];
      c. R[4] = R[4] ^ R[8];
      d. R[5] = R[5] ^ R[8];
      e. R[6] = R[6] ^ R[8];
      f. R = Trunc8[R].
    4. Return R[0].
  \end{lstlisting}
\end{lrbox}

\newsavebox{\RcCry}
\begin{lrbox}{\RcCry}
  \begin{lstlisting}[language=Cryptol, numbers=left, escapeinside={;}{;}]
    rc : [64] -> Bit
    rc t =  if (t % 255) == 0 then 1
            else rs !0 @0 where;\label{line:return};
              rs = [0b10000000] #;\label{line:init};
                   [if i <= t % 255 then;\label{line:compute};
                      (take'{8} ([(r@0) ^ (r@8),
                                   r@1, r@2, r@3, 
                                  (r@4) ^ (r@8),
                                  (r@5) ^ (r@8),
                                  (r@6) ^ (r@8),
                                   r@7,r@8])
                        where r = [0] # (rs @(i - 1)))
                    else rs @(i - 1);\label{line:stutter};
                   | i <- [1..254]];\label{line:alliters};
  \end{lstlisting}
\end{lrbox}

\newsavebox{\whilemethod}
\begin{lrbox}{\whilemethod}
  \begin{lstlisting}[language=Cryptol]
    while : {a} a -> (a -> Bit) -> (a -> a) -> a
    while state cond f = 
      if (cond state) then (while (f state) cond f)	
      else state
  \end{lstlisting}
\end{lrbox}

\newsavebox{\RcCryLib}
\begin{lrbox}{\RcCryLib}
  \begin{lstlisting}[language=Cryptol, numbers=left, escapeinside={;}{;}]
    rc : [64] -> Bit
    rc t =  if (t % 255) == 0 then 1 
            else (while {i = 1, R = 0b10000000};\label{line:state};
                        (&state -> state.i <= t % 255);\label{line:cond};
                        (&state -> {
                          i = state.i + 1,;\label{line:inc};
                          R = (take'{8} ([(r@0) ^ (r@8),
                                           r@1, r@2, r@3,
                                          (r@4) ^ (r@8),
                                          (r@5) ^ (r@8),
                                          (r@6) ^ (r@8),
                                           r@7, r@8]) 
                                where r = [0] # state.R)})
                 ).R @0;\label{line:libreturn};
  \end{lstlisting}
\end{lrbox}

\begin{figure}[t]
  \begin{center}
    \begin{tabular}{l}
      \usebox{\fipsRc}
    \end{tabular}
  \end{center}
  \caption{The \fips\ specification for \emph{rc}.}
  \label{fig:rc}
\end{figure}

Another common idiom in the \fips\ standard is to iteratively transform some state a fixed number of times to arrive at a final state.
The definition of the \emph{rc} function in \figref{fig:rc} is one such example.
The \emph{rc} function is a support function for the $\iota$ function in \keccak\ and is used to build a mask.

The \emph{rc} function checks the value of the input integer $t$, and returns the bit $0$ if it is a multiple of $255$.
If not, then Line 2 defines $R$ as the 8-bit value $128$, and Line 3 iteratively transforms $R$ a fixed number of times: $t\ \mathbf{mod}\ 255$.
The transform is defined on Line 3a through Line 3f.
Line 3a prepends a $0$ bit to $R$ so that it is now a 9-bit value. The array indexing refers to bit positions from left to right so $R[0]$ is the prepended left-most bit and $R[8]$ is the right-most bit. Each transform on $R$ prepends the 0 (Line 3a), does an exclusive-or with specific bit positions (Line 3b through Line 3e), and then takes the left-most 8 bits for the next round effectively dropping the right-most bit (Line 3f).
It returns the $R[0]$ at the end.

The \cryptol\ equivalent for \emph{rc} in \figref{fig:rccry}(a) is considerably more complex with the list comprehensions.
Returning $1$ when the input is a multiple of $255$ is straightforward.
It is when that is not the case that it is less obvious.
The list comprehension builds a list of $255$ 8-bit values.
\lineref{line:return} returns the bit at index 0 ($@0$) from the last 8-bit value in the array ($!0$).
The computation of the list of 8-bit values follows.

\lineref{line:init} is the first entry in the list and it is the initial value of $R$.
The following entries depend on the value of the input $t$ and the value of $i$ that comes from the list used in the list comprehension on \lineref{line:alliters}.
\lineref{line:compute} checks if $i$ is within the number of iterations, and if it is, then the new entry in the list is computed from the previous entry as defined by the transform in \figref{fig:rc}.
\lineref{line:stutter} is the case for when $i$ is not within the number of iterations.
Here the previous value in the list is copied to stutter the last value of $R$. 
The \cryptol\ specification creates all $255$ entries in the list regardless of the input value $t$ except for the case when $t$ is a multiple of 255.
As before, \figref{fig:rccry}(a) is not altogether unreadable, but it can be improved considerably.

\figref{fig:rccry}(b) is a recursive definition of a \emph{while-loop}.
It takes as input a state, a predicate on the state, \emph{cond}, that is true when it should loop and false otherwise, and a function, $f$, that computes a new state from the old state on each iteration.
The definition makes recursive calls to the while-method until the state predicate is false which point it returns the current value of \emph{state}.

\begin{figure}[h]
  \begin{center}
    \begin{tabular}{l}
      \usebox{\RcCry} \\ \\
      \multicolumn{1}{c}{(a)} \\ \\
      \usebox{\whilemethod} \\ \\
      \multicolumn{1}{c}{(b)} \\ \\ 
      \usebox{\RcCryLib} \\ \\
      \multicolumn{1}{c}{(c)}
    \end{tabular}
  \end{center}
  \caption{The \emph{rc} function in \cryptol. (b) The \cryptol\ with list comprehensions. (b) The recursive while-method definition. (c) The \cryptol\ with the while-method.}
  \label{fig:rccry}
\end{figure}

\figref{fig:rccry}(c) is the definition of \emph{rc} using the while-method in \figref{fig:rccry}(b).
\lineref{line:state} defines the state for the while loop to be both an integer $i$ for a counter and the 8-bit value $R$.
\lineref{line:cond} is the looping condition defined on the state.
The body of the loop not only computes the new value of $R$ but \lineref{line:inc} increments $i$ to track the loop iteration.
The return value is given on \lineref{line:libreturn} as the $0^\mathrm{th}$ bit from the last value of $R$.
Here, the only computes $R$ the number of times required by $t$ which differs significantly from the definition without the while-method.
As with the for-method, the while-method reduces the gap between how things are defined in \fips\ and in \cryptol.
